Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-05-31 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite2d-paper: para on GOcean infrastructure

\affil[1]{Science and Technology Facilities Council, Daresbury Laboratory, UK}
\affil[2]{NVIDIA}
\affil[3]{National Oceanography Centre, Liverpool, UK}
\affil[4]{University of Manchester, Manchester, UK}

@@

The challenge presented to the developers of scientific software by
the drive towards Exascale computing is considerable. With power
consumption becoming the overriding design constraint, CPU clock
speeds are falling and the complex, multi-purpose compute core is
being replaced by multiple, simpler cores. This philosophy can be seen
at work in the rise of so-called accelerator-based machines in the Top
500 List~\citep{top500} of supercomputers: four of the top-ten machines
in the June 2016 list make use of many-core processors (Intel Xeon
Phi, NVIDIA GPU or NRCPC SW26010). Two of the remaining six machines
are IBM BlueGene/Qs, the CPU of which has hardware support for running
64 threads.

@@

In a previous paper~\citep{shallow_psykal} we introduced a possible
approach to tackling this problem which we term {PS}y{KA}l (discussed
below). In that work we considered the implications for serial
performance of the extensive code re-structuring required by the
approach when applied to the `Shallow' shallow-water model
(https://puma.nerc.ac.uk/trac/GOcean). We found that although the
re-structuring did initially incur a sizeable performance penalty, it
was possible to transform the resulting code to recover performance
(for a variety of CPU/compiler combinations) while obeying the
{PS}y{KA}l separation of concerns. In this work we move to looking at
portable parallel performance within the {PS}y{KA}l approach.

@@

The PSy layer sits in-between the Algorithm and Kernel layers and its
functional role is to link the algorithm calls to the associated
kernel subroutines. As the Algorithm layer works on logically global
fields and Kernel layer works on local fields, the PSy layer is
responsible for iterating over columns. It is also responsible for
including any distributed-memory operations resulting from the
decomposition of the simulation domain, such as halo swaps and
reductions.

As the PSy layer iterates over columns, the single-core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be exploited and optimised. The PSy layer can
therefore be tailored for a particular hardware (such as multi-core,
many-core, GPGPUs, or some combination thereof) and software (such as
compiler, operating system, MPI implementation) configuration with no
change to the Algorithm or Kernel layer code. This approach therefore
offers the potential for portable performance. In this work we apply
optimisations to the PSy layer manually. The development of a tool to
automate this process will be the subject of a future paper.

@@

where $\zeta$ and $\vec{U}$ represent the sea-surface height and
horizontal velocity vector, respectively. $h$ is the total water
depth, $\Omega$ is the Earth rotation velocity vector.  $g$ is the
acceleration due to gravity. $\nu$ is the kinematic viscosity coefficient.

@@

@@

\begin{figure}
\centering
\begin{verbatim}
DO istep = 1, nsteps
  call invoke( continuity(ssha_t, ...), &
               momentum_u(ua, un, ...), &
               ...,                     &
               next_sshu(sshn_u, ...),  &
               next_sshv(sshn_v, ...) )
  call model_write(istep, sshn_t, un, vn)
END DO
\end{verbatim}
\caption{A schematic of the top-level of the {PS}y{KA}l version of the
 NEMOLite2D code.}
\label{FIG_psykal_nemolite2d_structure}
\end{figure}

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The compiler flags used in this work.}
\label{TABLE_compiler_flags}
\centering
\begin{tabular}{l|l|l}
\hline
Compiler  &  Flags                                    & Notes \\
\hline
Gnu       & -Ofast -mtune=native -finline-limit=50000 &   \\
Intel     & -O3 -fast -fno-inline-factor -xHost       &   \\
Cray      & -O3 -O ipa5 -h wp                         &   \\
PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 & \\
\hline
\end{tabular}
\end{table}

@@

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the NEMOLite2D kernels contain conditional statements. These
statements are there to check whether \textit{e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the \texttt{safe\_address}
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

>>>>

\affil[1]{Science and Technology Facilities Council, Daresbury Laboratory, UK}
\affil[2]{NVIDIA Corporation}
\affil[3]{National Oceanography Centre, Liverpool, UK}
\affil[4]{University of Manchester, Manchester, UK}

@@

The challenge presented to the developers of scientific software by
the drive towards Exascale computing is considerable. With power
consumption becoming the overriding design constraint, CPU clock
speeds are falling and the complex, multi-purpose compute core is
being replaced by multiple, simpler cores. This philosophy can be seen
at work in the rise of so-called accelerator-based machines in the Top
500 List of supercomputers (http://www.top500.org/): six of the
top-ten machines in the November 2016 list make use of many-core
processors (Intel Xeon Phi, NVIDIA GPU or NRCPC SW26010). Two of the
remaining four machines are IBM BlueGene/Qs, the CPU of which has
hardware support for running 64 threads.

@@

In a previous paper~\citep{shallow_psykal} we introduced a possible
approach to tackling this problem which we term {PS}y{KA}l (discussed
below). In that work we considered the implications for serial
performance of the extensive code re-structuring required by the
approach when applied to the `Shallow' shallow-water model
(https://puma.nerc.ac.uk/trac/GOcean). We found that although the
re-structuring did initially incur a sizeable performance penalty, it
was possible to transform the resulting code to recover performance
(for a variety of CPU/compiler combinations) while obeying the
{PS}y{KA}l separation of concerns. In this work we move to looking at
portable \textit{parallel} performance within the {PS}y{KA}l approach.

@@

The PSy layer sits in-between the Algorithm and Kernel layers and its
functional role is to link the algorithm calls to the associated
kernel subroutines. As the Algorithm layer works on logically global
fields and the Kernel layer works on local fields, the PSy layer is
responsible for iterating over columns. It is also responsible for
including any distributed-memory operations resulting from the
decomposition of the simulation domain, such as halo swaps and
reductions.

As the PSy layer iterates over columns, the single-core performance
can be optimised by applying transformations such as loop fusion and
loop blocking. Additionally, the potential parallelism within this
iteration space can also be exploited and optimised. The PSy layer can
therefore be tailored for a particular hardware (such as multi-core,
many-core, GPUs, or some combination thereof) and software (such as
compiler, operating system, MPI library, \textit{etc}.) configuration
with no change to the Algorithm or Kernel layer code. This approach
therefore offers the potential for portable performance. In this work
we apply optimisations to the PSy layer manually. The development of a
tool to automate this process will be the subject of a future paper.

@@

where $\zeta$ and $\vec{U}$ represent the sea-surface height and
horizontal velocity vector, respectively. $h$ is the total water
depth, $\Omega$ is the Earth rotation velocity vector.  $g$ is the
acceleration due to gravity and $\nu$ is the kinematic viscosity coefficient.

@@

In the Algorithm layer, fields (and grids) are treated as logically
global objects. Therefore, as part of creating the {PS}y{KA}l version
of NEMOLite2D, we represent fields with derived types instead of
arrays in this layer. These types then hold information about the
associated mesh and the extents of 'internal' and 'halo' regions as
well as the data arrays themselves. This frees the natural scientist
from having to consider these issues and allows for a certain degree
of flexibility in the actual implementation (e.g. padding for
alignment or increasing array extent to allow for other
optimisations). The support for this is implemented as a library
(which we term the GOcean Infrastructure) and is common to the
{PS}y{KA}l versions of both NEMOLite2D and Shallow.

@@

\begin{figure}
\centering
\begin{verbatim}
DO istep = 1, nsteps
  call invoke( continuity(ssha_t, ...), &
               momentum_u(ua, un, ...), &
               ...,                     &
               next_sshu(sshn_u, ...),  &
               next_sshv(sshn_v, ...) )
  call model_write(istep, sshn_t, un, vn)
END DO
\end{verbatim}
\caption{A schematic of the top-level of the {PS}y{KA}l version of the
  NEMOLite2D code. The kernels listed as arguments to the \texttt{invoke}
call specify the operations to be performed.}
\label{FIG_psykal_nemolite2d_structure}
\end{figure}

@@

\begin{table}[!t]
% increase table row spacing, adjust to taste
\renewcommand{\arraystretch}{1.3}
\caption{The compiler flags used in this work.}
\label{TABLE_compiler_flags}
\centering
\begin{tabular}{l|l|l}
\hline
Compiler  &  Flags                                                       \\
\hline
Gnu       & -Ofast -mtune=native -finline-limit=50000                    \\
Intel     & -O3 -fast -fno-inline-factor -xHost                          \\
Cray      & -O3 -O ipa5 -h wp                                            \\
PGI       & -acc -ta=tesla,cc35,nordc -Mcuda=maxregcount:80,loadcache:L1 \\
\hline
\end{tabular}
\end{table}

@@

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the NEMOLite2D kernels contain conditional statements. These
statements are there to check whether \textit{e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the \texttt{safe\_address}
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).
